On hyperparameter tuning through Lipschitz global optimization

### 2024-03-29 2024/ITMO paper/sn-article.tex
remove string numbers
submitted paper

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
\documentclass[lineno,iicol]{sn-jnl}
%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

@@

@@

\affil[2]{
%\orgdiv{Department},
\orgname{ITMO University}, \orgaddress{\street{Lomonosova St. 9}, \city{St. Petersburg}, \postcode{191002}, 
%\state{State},
\country{Russia}}}

@@

Currently, complex scientific and/or industrial problems are most often solved by numerical methods. Most of such methods, for example, machine learning methods, have hyperparameters, and the choice of specific values of these hyperparameters can determine the efficiency (in a given metric) of the obtained solution under certain constraints on the computation time \citep{Hutter2019,nikitin2021}. The fact that in many cases the difference in the efficiency of the solution obtained with different values of hyperparameters can be quite significant, has resulted in the emergence of the hyperparameter optimization (HPO) problem, as well as the research of approaches to its solution that are implemented in the frameworks for automatic selection of hyperparameters \citep{Tune,optuna,hyperopt,Sherpa}. From the mathematical point of view, the HPO problem corresponds to a global optimization problem with fixed bounds on the variation of hyperparameters.  In certain cases, some of the hyperparameters may be categorical, i.e., they can only assume values from a certain, usually small, number of variants, which necessitates the use of optimization methods capable of solving problems with discrete parameters.

@@

@@

The iOpt framework contains an implementation of Lipschitz optimization algorithms, which are an extension of the information-statistical global search algorithm described in detail in \citet{Strongin2000,Sergeyev2013}. In particular, the methods have been adapted to solve problems with partially integer parameters, which are very common in HPO problems. The experimental results presented in Section \ref{sec_exp} show that the iOpt framework is not inferior in HPO problems to well-known frameworks having the same purpose (such as Optuna \citep{optuna} and HyperOpt \citep{hyperopt}).

@@

\subsection{Results on live hyperparameters tuning}

@@

To obtain a good solution, the methods from Optuna and Hyperopt should be restarted several times. If one focuses on the total operating time, the performance gain of iOpt is even greater. As an example, Figure \ref{fig8} shows a comparison of the hyperparameter tuning times for iOpt, Optuna and Hyperopt on several datasets and different classification models. For iOpt, the time is shown as a blue dot. For Optuna and Hyperopt, the box plots are shown.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{fig_time}
\caption{Time of hyperparameters tuning.} \label{fig8}
\end{figure*}

\subsection{Results on surrogate models}

@@

>>>>

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
\documentclass[iicol]{sn-jnl}
%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

@@

\author[2]{\fnm{Lubov} \sur{Yamschikova}}

\author[2]{\fnm{Nikolay O.} \sur{Nikitin}}

@@

\affil[2]{
%\orgdiv{Department},
\orgname{ITMO University}, \orgaddress{\street{Kronverksky Ave. 49, bldg. A}, \city{St. Petersburg}, \postcode{197101}, 
%\state{State},
\country{Russia}}}

@@

Currently, complex scientific and/or industrial problems are most often solved by numerical methods. Most of such methods, for example, machine learning methods, have hyperparameters, and the choice of specific values of these hyperparameters can determine the efficiency (in a given metric) of the obtained solution under certain constraints on the computation time \citep{Hutter2019,nikitin2022automated}. The fact that in many cases the difference in the efficiency of the solution obtained with different values of hyperparameters can be quite significant, has resulted in the emergence of the hyperparameter optimization (HPO) problem, as well as the research of approaches to its solution that are implemented in the frameworks for automatic selection of hyperparameters \citep{Tune,optuna,hyperopt,Sherpa}. From the mathematical point of view, the HPO problem corresponds to a global optimization problem with fixed bounds on the variation of hyperparameters.  In certain cases, some of the hyperparameters may be categorical, i.e., they can only assume values from a certain, usually small, number of variants, which necessitates the use of optimization methods capable of solving problems with discrete parameters.

@@

%Nikitin
Due to the high complexity of many HPO tasks, the meta-learning techniques are frequently used. For example, deep learning models \citep{doke2021survey} are usually too computationally expensive to tune the hyperparameters directly. In resent studies, large language models are used as powerful meta-models \citep{zhang2023using}. However, the deficiency of meta-learning based approach as not sufficient for many machine learning applications \citep{doke2021survey}.

@@

The iOpt framework contains an implementation of Lipschitz optimization algorithms, which are an extension of the information-statistical global search algorithm described in detail in \citet{Strongin2000,Sergeyev2013}. In particular, the methods have been adapted to solve problems with partially integer parameters, which are very common in HPO problems. The experimental results presented in Section~\ref{sec_exp} show that the iOpt framework is not inferior in HPO problems to well-known frameworks having the same purpose (such as Optuna~\citep{optuna} and HyperOpt~\citep{hyperopt}).

@@

\subsection{Results on ML models}

@@

To obtain a good solution, the methods from Optuna and Hyperopt should be restarted several times. If one focuses on the total operating time, the performance gain of iOpt is even greater. As an example, Fig.~\ref{fig8} shows a comparison of the hyperparameter tuning times for iOpt, Optuna and Hyperopt on several datasets and different classification models. For iOpt, the time is shown as a blue dot. For Optuna and Hyperopt, the box plots are shown.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{fig_time}
\caption{Time of hyperparameters tuning.} \label{fig8}
\end{figure*}

\subsection{Results on surrogate HPO benchmark}

@@

\subsection{Results on AutoML-designed pipelines}

In many cases, HPO is considered not as an independent machine learning task but as a part of automated machine learning workflow (see, e.g., \cite{filippou2023structure}). The AutoML task is commonly formulated as CASH -- Combined Algorithms Selection and Hyperparameter Optimization \citep{kim2022survey}. Since state-of-the-art AutoML tends to produce complicated ensemble-based modeling pipelines, the efficiency of HPO can differ for AutoML-generated models and static tasks in benchmarks.

In the paper, we choose the FEDOT AutoML framework \citep{nikitin2022automated} as a test bench since it allows the production of diverse pipelines with different complexities. The performance of HPO for the tuning of ensemble-based models was compared to the results of optimization algorithms based on Optuna and Hyperopt frameworks. Optimizers were compared on ensemble machine-learning pipelines designed by the FEDOT framework. In this case, the pipelines are ensembles of ML algorithms assembled together using stacking. Such compositions of algorithms can be represented in the form of directed acyclic graphs.

For multi-node pipeline, the optimization strategy can be implemented in different ways:
\begin{enumerate}
    \item Optimization of hyperparameters in each node sequentially, maximizing the metric for the parameters of the entire graph.
    \item Optimization of all hyperparameters of the model simultaneously.
\end{enumerate}

In this experiment, the second approach was used; the search space was formed from all the hyperparameters of each of the vertices of the graph.

The algorithms were compared for regression task. %two machine learning tasks: regression and classification. 
The optimizers were run for a limited number of iterations (20 and 100). Each of the experiments was conducted with a different set of test models, datasets, and algorithms and was run for 30 times. Based on the results of the experiments, the average increase in the target metric on test data was compared, that is, the difference in metrics for models with the default hyperparameters and with the hyperparameters obtained after optimization.

%\subsubsection{Regression}
For the regression problem, three directed cyclic graphs were chosen to represent the different pipeline structures for the regression problem. These graphs differed in the complexity of their structure and the number of hyperparameters for optimization. The following types of graphs were taken as test ones: linear, branched, and complex (see Fig.~\ref{fig_reg_pipe}).

\begin{figure}
\center{\includegraphics[width=1.0\linewidth]{regression_pipe.png} }
\caption{Test graphs for regression task: a) linear pipeline, where \textit{knnreg} -- k-nearest neighbors algorithm, \textit{ridge} -- linear regression with L2 regularization; b) branched pipeline, where \textit{svr} -- support vector machine, \textit{rfr} -- random forest algorithm; c) complex pipeline, where \textit{lasso} -- linear regression with L1 regularization.}\label{fig_reg_pipe}
\end{figure}

Optimization of the hyperparameters for each pipeline was conducted for three data sets:
\begin{enumerate}
    \item California Housing Prices (CHP) -- data contains information from the 1990 California Census and provides average home prices for various quarters. The average cost of houses is the target variable.
    \item Delta Ailerons (DA) -- this data set is derived from the F16 aircraft aileron control problem. The attributes describe the state of the aircraft, and the goal is to predict the control input on the aircraft's ailerons.
    \item Pol -- a data set with forty-eight features describing a telecommunications problem.
\end{enumerate}

\begin{table*}
\centering
\caption{Metric improvement for regression task, where L -- linear pipeline, B -- branched pipeline, C -- complex pipeline.}
\label{tab:automl_reg}
\begin{tabular}{llccccc} \hline
Dataset & № iters & Test graph& iOpt & Hyperopt(tpe) & Optuna(tpe) & p-value\\ \hline
\multirow{6}{*}{CHP} & \multirow{3}{*}{20} & L& 1.2 & 1.2$\pm$0.1& 0.9$\pm$0.6& 0.9048 \\
 &  & B& 2.2& 2.2$\pm$0.3& \textbf{2.5$\pm$0.3}& 0.0001\\
 &  & C& \textbf{4.6} & 3.2$\pm$0.4& 3.2$\pm$0.6& 0.0000 \\
 & \multirow{3}{*}{100} & L
& 1.1& 1.1$\pm$0.0& 1.1$\pm$0.2& -\\
 &  & B
& 2.0& 2.1$\pm$0.0& \textbf{2.5$\pm$0.3}& 0.0000\\
 &  & C& \textbf{4.6} & 4.1$\pm$0.3& 4.0$\pm$0.6& 0.0000 \\ \hline
\multirow{6}{*}{DA} & \multirow{3}{*}{20} & L 
& \textbf{4.2} & 0.5$\pm$1.8& 0.5$\pm$1.1& 0.0000 \\ 
 &  & B
& \textbf{1.7}& 1.4$\pm$0.1& 1.4$\pm$0.1& 0.0000\\
 &  & C& \textbf{62.3} & 33.5$\pm$31.4& 11.9$\pm$23.6& 0.0001 \\
 & \multirow{3}{*}{100} & L
& 4.2 & \textbf{6.3$\pm$2.1}& 6.1$\pm$2.3& 0.0005 \\
 &  & 
B& \textbf{1.7}& 1.5$\pm$0.2& 1.5$\pm$0.2& 0.0000\\
 &  & C& 62.3 & \textbf{63.3$\pm$1.4}& 50.6$\pm$24.3& 0.0065 \\ \hline
\multirow{6}{*}{Pol} & \multirow{3}{*}{20} & 
L
& \textbf{1.2} & 0.8$\pm$0.3& 0.4$\pm$0.6& 0.0000 \\
 &  & 
B& \textbf{19.0}& 1.5$\pm$2.2& 3.7$\pm$0.35.4& 0.0006\\
 &  & C& 0.0 & 0.0$\pm$0.0& 0.0$\pm$0.0& -  \\ 
 & \multirow{3}{*}{100} & L& \textbf{1.2} & \textbf{1.2$\pm$0.0}& 1.0$\pm$0.5& 0.0000 \\
 &  &  B& \textbf{19.0}& 17.8$\pm$2.3& 13.9$\pm$5.3& 0.0000\\
 &  & C& 0.0 & 0.0$\pm$0.0& 0.0$\pm$0.0& - \\ \hline
\end{tabular}
\end{table*}

The results and their statistical significance (Friedman and Nemenyi tests) are shown in Table~\ref{tab:automl_reg}.
The experiments confirmed that the performance of iOpt in the HPO problem for ensemble models is comparable to Optuna, however, the undisputed advantage of the iOpt algorithm is its deterministic nature.
